---
title: "13 Build Final SIF"
format: gfm
execute:
  eval: false
filters:
  - sorting-hat
  - ripper
extensions:
  ripper:
    output-name: "build_final_sif"
    output-dir: "../../src/rse_workbench"
  sorting-hat:
    keep: python
---

This notebook builds the final read-only SIF from the validated platform definition.

```{python}
#| sorting-hat: remove
import submitit
from rse_workbench.helpers import show_submitit_job, spython_exec, show_spython
from rse_workbench.build_install import build_container
```

Build: 

```{python}
#| sorting-hat: remove
executor = submitit.AutoExecutor(folder=str(project_contract.submitit_dir / "%j"))
executor.update_parameters(
    timeout_min=project_contract.slurm_build_hours * 60,
    slurm_partition=project_contract.slurm_build_partition,
    mem_gb=float(project_contract.slurm_build_mem.rstrip("G")),
    cpus_per_task=project_contract.slurm_build_cpus,
    slurm_job_name=f"build-{project_contract.project_name}-container",
    mail_user=project_contract.slurm_email,
)
```

```{python}
job = executor.submit(build_container, project_contract.singularity_def, project_contract.singularity_image)
job.job_id
```

```{python}
result = show_submitit_job(job)
if result is not None:
    print(result)
```

Install Spack and System Dependencies:

```{python}
job = executor.submit(
    install_in_container,
    project_dir=project_contract.project_dir,
    SINGULARITY_IMAGE=project_contract.singularity_image,
    runtime_command=project_contract.spack_install_command,
)
job.job_id
```

```{python}
result = show_submitit_job(job)
if result is not None:
    print(''.join(result['message']))
```

Install Data Science Languages and Packages:

```{python}
job = executor.submit(
    install_in_container,
    project_dir=project_contract.project_dir,
    SINGULARITY_IMAGE=project_contract.singularity_image,
    runtime_command=project_contract.language_install_command,
)
job.job_id
```

```{python}
result = show_submitit_job(job)
if result is not None:
    print(''.join(result['message']))
```

This section verifies that each layer of the environment is functioning correctly **independently** before testing that the layers communicate correctly.

The architecture is expected to look like:

```text
Container
    ↓
Spack
    ↓
R / Rscript
    ↓
rv-managed R package library
    ↓
Quarto
    ↓
Project render
```

At each stage we test a single contract before moving to the next.

---

## 1. Verify the container runtime

The container should provide the basic executables required for the project.

```{python}
command = r"""
set -euo pipefail
echo "============================================================"
echo "1. Base container tools (before activation)"
echo "============================================================"

command -v spack
command -v rv
command -v uv
command -v quarto
quarto --version
echo

echo "============================================================"
echo "2. Activate project"
echo "============================================================"

cd /work
source /work/env/activate.sh
echo

echo "============================================================"
echo "3. Tool discovery after activation"
echo "============================================================"

command -v spack
command -v R
command -v Rscript
command -v rv
command -v quarto
echo

echo "============================================================"
echo "4. Spack"
echo "============================================================"

spack env status
spack find r
echo

echo "============================================================"
echo "5. rv"
echo "============================================================"

rv cache
rv sync
echo

echo "============================================================"
echo "6. R"
echo "============================================================"

Rscript -e '
cat("R.home():\n")
print(R.home())
cat("\n.libPaths():\n")
print(.libPaths())

pkgs <- c(
  "rlang",
  "knitr",
  "rmarkdown"
)

for (pkg in pkgs) {
  cat("\n=====================================\n")
  cat(pkg, "\n")
  cat("=====================================\n")
  cat("find.package():\n")
  print(find.package(pkg, quiet = TRUE))
  cat("\nrequireNamespace():\n")
  print(requireNamespace(pkg, quietly = FALSE))
  cat("\nlibrary():\n")
  try(library(pkg, character.only = TRUE))
  cat("\npackageVersion():\n")
  print(packageVersion(pkg))
}
'
echo

echo "============================================================"
echo "7. Quarto discovery"
echo "============================================================"

export QUARTO_R="$(command -v Rscript)"
echo "QUARTO_R=${QUARTO_R}"
quarto check knitr --log-level DEBUG
echo

echo "============================================================"
echo "8. Minimal Quarto render"
echo "============================================================"

tmpdir="$(mktemp -d)"

cat > "${tmpdir}/test.qmd" <<'QMD'

---
title: "Smoke Test"
format: html
---

```{r}
sessionInfo()
library(rlang)
library(knitr)
library(rmarkdown)

1 + 1
```
QMD

quarto render "${tmpdir}/test.qmd" \
    --output-dir "${tmpdir}/out"

echo
echo "Output:"
ls -lah "${tmpdir}/out"

echo
echo "DONE"
"""

```

Expected:

- every executable is found
- Quarto prints its version

---

## 2. Verify that Spack owns the R installation

Activate the project environment and inspect the R installation.

```{python}
show_spython(
    spython_exec(
        command=r'''
cd /work

source /work/env/activate.sh

spack env status
spack find r

Rscript -e '
cat("R.home():\n")
print(R.home())

cat("\n.libPaths():\n")
print(.libPaths())
'
''',
        image=SINGULARITY_IMAGE,
        options=[
            "--home", f"{CONTAINER_HOME}:/home/rse",
            "--bind", f"{CONTAINER_TMP}:/tmp",
        ],
        bind_project=PROJECT_DIR,
    ),
)
```

Expected:

- the project Spack environment is active
- `R.home()` points inside `.spack-env`
- `.libPaths()` contains

```
/work/.rv/library
/work/.spack-env/.../R/library
```

---

## 3. Verify that rv owns the project R packages

Synchronize the project packages and verify that they can be loaded.

```{python}
show_spython(
    spython_exec(
        command=r'''
rv sync

Rscript -e '
cat(".libPaths():\n")
print(.libPaths())

pkgs <- c("rlang", "knitr", "rmarkdown")

for (pkg in pkgs) {
  cat("\n=============================\n")
  cat(pkg, "\n")
  cat("=============================\n")

  print(find.package(pkg, quiet = TRUE))
  print(requireNamespace(pkg, quietly = FALSE))
}
'
''',
        image=SINGULARITY_IMAGE,
        options=[
            "--home", f"{CONTAINER_HOME}:/home/rse",
            "--bind", f"{CONTAINER_TMP}:/tmp",
        ],
        bind_project=PROJECT_DIR,
    ),
)
```

Expected:

```
TRUE
TRUE
TRUE
```

for all three packages.

---

## 4. Verify that Quarto discovers the same R installation

Explicitly tell Quarto which R executable to use.

```bash
export QUARTO_R="$(command -v Rscript)"

quarto check knitr --log-level DEBUG
```

Expected:

- Quarto reports the same R installation
- knitr is detected
- rmarkdown is detected

---

## 5. Verify that Quarto can render a minimal document

Create a temporary document that exercises the R engine.

```bash
tmp="$(mktemp -d)"

cat > "$tmp/test.qmd" <<'EOF'
---
title: "Quarto Smoke Test"
format: html
---

```{r}
library(knitr)
library(rmarkdown)

1 + 1
```
EOF

quarto render "$tmp/test.qmd" --output-dir "$tmp/out"

ls -lah "$tmp/out"
```

Expected:

- rendering succeeds
- HTML output is produced

---

## Interpretation

The expected progression is

```
Container
    ✓

Spack
    ✓

R
    ✓

rv
    ✓

Quarto
    ✓

Render
    ✓
```

The first failing step identifies the boundary where integration has broken.

There is no benefit to debugging components beyond the first failed contract.


```{python}
show_spython(
    spython_exec(
        command=r'''
set -euo pipefail
echo "PATH=$PATH"
cd /work
source /work/env/activate.sh

quarto render dev
''',
        image=SINGULARITY_IMAGE,
        options=[
            "--home", f"{CONTAINER_HOME}:/home/rse",
            "--bind", f"{CONTAINER_TMP}:/tmp",
        ],
        bind_project=PROJECT_DIR,
    ),
)   
```